---
title: "Data Management with R: An Inroduction to dplyr"
author: "Jubayer Hossain"
date: "21 February, 2021"
categories: ["R Programming"]
summary: dplyr is a new package which provides a set of tools for efficiently manipulating datasets in R.
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment=NA, fig.align = "center", fig.width = 8, fig.height = 4)
```

## What is `dplyr`? {.t}
- The `dplyr` package was developed by Hadley Wickham of RStudio. 
- `dplyr` is a new package which provides a set of tools for efficiently manipulating datasets in R.
- `dplyr` is the next iteration of `plyr` , focussing on only data frames. 
- With `dplyr` , anything you can do to a local data frame you can also do to a remote database table.

## `dplyr` Functionality {.t}
- Five basic verbs: `filter`, `select`, `arrange`, `mutate`, `summarise` and `groub_by`
- Can work with data stored in databases and data tables
- Joins: inner join, left join, semi-join, anti-join
- Window functions for calculating ranking, offsets, and more
- Better than plyr if you’re only working with data frames (though it doesn’t yet duplicate all of the plyr functionality)

## Why `dplyr`? {.t}
- Great for data exploration and transformation
- Intuitive to write and easy to read, especially when using the “chaining” syntax (covered below)
- Fast on data frames

## `dplyr` Grammar{.t}
- `select`: return a subset of the column of a data franme, using a flexible notation. 
- `filter`: extract a subset of rows from a data frame based on logical conditions.
- `arrange`: reorder rows of data frame 
- `mutate`: add new variables/columns or transform existing variables. 
- `summarise/summarize`: generate summary statistics of different variables in the data frame, possibly within strata. 
- `%>%` "pipe" operator used to connect multiple verb actions together into a pipeline. 

## Installing `dplyr` {.t}
```r
install.packages('dplyr')
```

## Loading Data: The Gapminder Dataset{.t}
```r
# Install gapminder dataset 
install.packages('gapminder')
```
```{r}
# Load gapminder dataset 
library(gapminder)
```

## Exploring the Gapminder Dataset {.t}
```{r}
# Examine first few rows 
head(gapminder)
```
```{r}
# Examine last few rows 
tail(gapminder)
```

```{r}
# Dimensions 
dim(gapminder)

# Colnames 
names(gapminder)
```
```{r}
# Data Structures
str(gapminder)
```
```{r}
# Summary 
summary(gapminder)
```

## Command Structure (for all dplyr verbs){.t}
- first argument is a data frame
- return value is a data frame
- nothing is modified in place
- Note: dplyr generally does not preserve row names

## Load `dplyr` Package 
```{r}
# Load dplyr 
library(dplyr)
```

## `select`: Pick Single Column by Name{.t}
```{r}
select(gapminder, country)
```
## `select`: Pick Multiple Columns by Name{.t}
```{r}
select(gapminder, country, continent, year)
```


## `select`: Removing Single Column{.t}
```{r}
select(gapminder, - gdpPercap)
```
## `select`: Removing Multiple Columns 
```{r}
select(gapminder, -c(pop,  gdpPercap))
```
## `select`: Select Column Using : (Range){.t}
```{r}
select(gapminder, country:year)
```
## `select`: Select Single Column Using "`contains `"{.t}
```{r}
select(gapminder, contains("c"))
```

## `select`: Select Multiple Columns Using "`contains `"{.t}
```{r}
select(gapminder, contains("c"), contains('g'))
```

## `select`: Select Column Using "`starts_with`" {.t}
```{r}
select(gapminder, starts_with('c'))
```
## `select`: Select Column Using "`ends_with`"{.t}
```{r}
select(gapminder, ends_with('p'))
```
## Chaining Method: The Pipe(%>%) Operator{.t}
```{r}
gapminder %>% 
  select(country, continent, year) %>% 
  head() 
```
## The Count Verb{.t}
```{r}
gapminder %>% 
  count()
```
## Count Variable{.t}
```{r}
gapminder %>% 
  count(country)
```

## Count and Sort{.t}
```{r}
gapminder %>% 
  count(country, sort = TRUE)
```

## Count Population{.t}
```{r}
gapminder %>% 
  select(country, pop) %>% 
  count(country, wt = pop, sort = TRUE)

```


## filter: Equality(`"=="`){.t}
```{r}
filter(gapminder, country == "Bangladesh")
```
## filter: Inequality(`"!="`){.t}
```{r}
filter(gapminder, country != "Bangladesh")
```

## filter: Greater(`">"`){.t}
```{r}
filter(gapminder, gdpPercap > 800)
```

## filter: Greater or Equal(`">="`){.t}
```{r}
filter(gapminder, gdpPercap >= 800)
```

## filter: Less(`"<"`){.t}
```{r}
filter(gapminder, gdpPercap < 800)
```

## filter: Less or Equal(`"<="`){.t}
```{r}
filter(gapminder, gdpPercap <= 800)
```

## filter: Logical AND(`"&"`){.t}
```{r}
filter(gapminder, country=="Bangladesh" & gdpPercap > 800)
```

## filter: Logical OR(`"|"`){.t}
```{r}
filter(gapminder, country == "Bangladesh" |  gdpPercap > 800)
```

## filter: The `"%in%"` Operator{.t}
```{r}
filter(gapminder, country %in% c("Bangladesh", "Australia"))
```

## `mutate`: Creating New Column{.t}
```{r}
gapminder %>% 
  mutate(gdp = gdpPercap * pop) %>% 
  head() 
```

```{r}
# GDP in Million 
gapminder %>% 
  mutate(gdp = gdpPercap * pop / 10^6) %>% 
  head() 
```

## `arrange` : Reorder Rows{.t}
```{r}
gapminder %>% 
  select(country, pop) %>% 
  arrange(pop) %>% 
  head() 
```

## `arrange` : Reorder Rows(Descending){.t}
```{r}
gapminder %>% 
  select(country, pop) %>% 
  # descending order 
  arrange(desc(pop)) %>% 
  head() 
```

## `group_by`: Grouping Data 
```{r}
gapminder %>% 
  group_by(continent) %>% 
  head()
```

## `group_by` and `summarize` {.t}
```{r}
gapminder %>% 
  summarize(total_pop = sum(pop))
```

## Aggregate and summarize{.t}
```{r}
gapminder %>% 
  summarize(total_population = sum(pop), 
            avg_gdppercap = mean(gdpPercap))
```

## `summarise`: Sum{.t} 
```{r}
gapminder %>% 
  group_by(continent) %>% 
  summarise(pop = sum(pop))
```

## `summarise`: Sum{.t} 
```{r}
gapminder %>% 
  group_by(continent) %>% 
  # In Million
  summarise(pop = sum(pop) / 10^6)
```

## `summarise`: Maximum{.t}
```{r}
gapminder %>% 
  group_by(continent) %>% 
  summarise(max_liexp = max(lifeExp))
```

## Summary Functions{.t}
```r
mean() 
sum() 
median() 
min() 
max() 
n()
```

## Aggregate within Groups{.t}
```{r}
gapminder %>% 
  group_by(continent) %>% 
  summarize(total_pop = sum(pop),
            avg_lifeexp = mean(lifeExp))
```

## The `top_n` Verb 
```{r}
gapminder %>% 
  group_by(continent) %>% 
  top_n(1, pop)
```

## `rename`: Renaming Column{.t}
```{r}
gapminder %>% 
  rename(population = pop) %>% 
  head() 
```
